---
title: "Perceptual Illusion"
subtitle: "Data Analysis"
author: "Fernando Millan Villalobos"
date: "February 2021"
output:
  html_document:
    code_folding: show
    echo: TRUE
    warning: FALSE
    message: FALSE
    highlight: pygments
    theme: paper
    df_print: kable
    toc: yes
    toc_depth: 4
    number_sections: yes
    toc_float: 
      collapsed: yes
      smooth_scroll: false
  pdf_document:
    toc: yes
  word_document:
    toc: yes
  allways_allow_html: yes
---

```{r, echo=FALSE}
# CONFIG
user_name <- "fernandomillanvillalobos" # your Git username (only needed if
# you want to deploy to GH pages)
project_name <- "2020-perceptual-illusion-master" # adapt!
package_date <- "2021-01-01" # date of the CRAN snapshot that
# the checkpoint package uses
r_version <- "4.0.3" # R-Version to use
options(Ncpus = 4) # use 4 cores for parallelized installation of packages
if (r_version != paste0(version$major, ".", version$minor)) {
  stop("ERROR: specified R version does not match currently used.")
}
```

# Notes

This report was generated on `r Sys.time()`. R version: `r paste0(version$major, ".", version$minor)` on `r version$platform`. For this report, CRAN packages as of `r package_date` were used.

...

## R-Script & data

The pre-processing and analysis of the data was conducted in the [R project for statistical computing](https://www.r-project.org/). The RMarkdown script used to generate this document and all the resulting data can be downloaded [under this link](http:fernandomillanvillalobs.github.io/2020-perceptual-illusion-master/). Through executing `main.Rmd`, the herein described process can be reproduced and this document can be generated. In the course of this, data from the folder `input` will be processed and results will be written to `output`. The html on-line version of the analysis can be accessed through this [link](https://fernandomillanvillalobos.github.io/2020-perceptual-illusion-master/).

## GitHub

The code for the herein described process can also be freely downloaded from [https://github.com/fernandomillanvillalobs/2020-perceptual-illusion-master](https://github.com/fernandomillanvillalobos/2020-perceptual-illusion-master).

## Data description of output files

#### `abc.csv` (Example)

| Attribute | Type    | Description |
|-----------|---------|-------------|
| a         | Numeric | ...         |
| b         | Numeric | ...         |
| c         | Numeric | ...         |

# Set up

```{r, echo=FALSE}
detach_all_packages <- function() {
  basic_packages_blank <-  c("stats",
                             "graphics",
                             "grDevices",
                             "utils",
                             "datasets",
                             "methods",
                             "base")
  basic_packages <- paste("package:", basic_packages_blank, sep = "")

  package_list <- search()[
    ifelse(unlist(gregexpr("package:", search())) == 1, TRUE, FALSE)]

  package_list <- setdiff(package_list, basic_packages)

  if (length(package_list) > 0)  for (package in package_list) {
    detach(package, character.only = TRUE, unload = TRUE)
    print(paste("package ", package, " detached", sep = ""))
  }
}

detach_all_packages()

# this allows multiple persons to use the same RMarkdown
# without adjusting the working directory by themselves all the time
source("scripts/csf.R")
path_to_wd <- csf() # if this - for some reason - does not work,
# replace with a hardcoded path, like so: "~/projects/rddj-template/analysis/"
if (is.null(path_to_wd) | !dir.exists(path_to_wd)) {
  print("WARNING: No working directory specified for current user")
} else {
  setwd(path_to_wd)
}

# suppress scientific notation
options(scipen = 999)

# unload global rstudioapi and knitr again to avoid conflicts with checkpoint
# this is only necessary if executed within RStudio
# outside of RStudio, namely in the knit.sh script, this causes RMarkdown
# rendering to fail, thus should not be executed there
if (Sys.getenv("RSTUDIO") == "1") {
  detach_all_packages()
}
```

## Define packages

```{r, echo=TRUE, message=FALSE, warning=FALSE}
# from https://mran.revolutionanalytics.com/web/packages/\
# checkpoint/vignettes/using-checkpoint-with-knitr.html
# if you don't need a package, remove it from here (commenting not sufficient)
# tidyverse: see https://blog.rstudio.org/2016/09/15/tidyverse-1-0-0/
cat("
library(rstudioapi)
library(tidyverse) # ggplot2, dplyr, tidyr, readr, purrr, tibble, magrittr, readxl
library(scales) # scales for ggplot2
library(jsonlite) # json
library(lintr) # code linting
library(sf) # spatial data handling
library(rmarkdown)
library(cowplot) # theme
library(extrafont) # fonts
library(janitor)", # names  
file = "manifest.R")
```

## Install packages

```{r, echo=TRUE, message=FALSE, warning=FALSE}
# if checkpoint is not yet installed, install it (for people using this
# system for the first time)
if (!require(checkpoint)) {
  if (!require(devtools)) {
    install.packages("devtools", repos = "http://cran.us.r-project.org")
    require(devtools)
  }
  devtools::install_github("RevolutionAnalytics/checkpoint",
                           ref = "v0.4.10", # could be adapted later,
                           # as of now (beginning of July 2017
                           # this is the current release on CRAN)
                           repos = "http://cran.us.r-project.org")
  require(checkpoint)
}
# nolint start
if (!dir.exists("~/.checkpoint")) {
  dir.create("~/.checkpoint")
}
# nolint end
# install packages for the specified CRAN snapshot date
checkpoint(snapshotDate = package_date,
           project = path_to_wd,
           verbose = T,
           scanForPackages = T,
           use.knitr = F,
           R.version = r_version)
rm(package_date)
```

## Load packages

```{r, echo=TRUE, message=FALSE, warning=FALSE}
source("manifest.R")
unlink("manifest.R")
sessionInfo()
```

## Load additional scripts

```{r, echo=TRUE, message=FALSE, warning=FALSE}
# if you want to outsource logic to other script files, see README for 
# further information
# Load all visualizations functions as separate scripts
knitr::read_chunk("scripts/dviz.supp.R")
source("scripts/dviz.supp.R")
knitr::read_chunk("scripts/themes.R")
source("scripts/themes.R")
knitr::read_chunk("scripts/plot_grid.R")
source("scripts/plot_grid.R")
knitr::read_chunk("scripts/align_legend.R")
source("scripts/align_legend.R")
knitr::read_chunk("scripts/label_log10.R")
source("scripts/label_log10.R")
```

# Data Wrangling

Importing datasets.

```{r importing}
# Read in data file
df.j <- read.table("input/j_ds.csv", sep = ",", header = TRUE, check.names = FALSE)

# Cleaning variables names
df.j <- janitor::clean_names(df.j)
```

Cleaning Julien's ds variables.

```{r recoding}
# Regrouping hours_sleep bottom and top categories
df.j$hours_sleep[df.j$hours_sleep == "<3"] <- "less_than_3"
df.j$hours_sleep[df.j$hours_sleep == ">8"] <- "more_than_8"

# Recoding training categories
df.j$photo_training[df.j$photo_training == "no training"] <- "no_training" 
df.j$carto_gis_training[df.j$carto_gis_training == "no training"] <- "no_training" 
df.j$carto_gis_training[df.j$carto_gis_training == "5"] <- "Professional" 

# Filling empty values for circle_below and circle_above as NAs
df.j$circle_below[df.j$circle_below == ""] <- NA 
df.j$circle_above[df.j$circle_above == ""] <- NA 
```

# Linting

The code in this RMarkdown is linted with the [lintr package](https://github.com/jimhester/lintr), which is based on the [tidyverse style guide](http://style.tidyverse.org/).

```{r echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
lintr::lint("main.Rmd", linters =
              lintr::with_defaults(
                commented_code_linter = NULL,
                trailing_whitespace_linter = NULL
                )
            )
# if you have additional scripts and want them to be linted too, add them here
# lintr::lint("scripts/my_script.R")
```
